%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgements

Like everything in life, this thesis was not possible without the help of many who deserve my respect and consideration.

First, I want to start by thanking my supervisors. To professor Rui Cortesão, my sincere thanks for giving me the opportunity to do a thesis in medical robotics and believing in me, even without a background in electrical engineering and robotics. It was a great learning experience. I also want to thank professor José Fonseca for being my co-supervisor and giving me some support when needed.

I want to thank the faculty, and specially the biomedical engineering group from the physics department. This area is amazing, very innovative and full of potential. We learned many interesting things. Special thanks to professor Carla Quintão for the work done on coordinating the course, and professor Hugo Gamboa for allowing me to work in his research group with such an awesome team. I learned a lot and made good friends. Thank you to the rest of the teachers with whom I learned, talked, discussed, laugh and received support: Cláudia Quaresma, Pedro Vieira, Ricardo Vigário, José Fonseca, Valentina Vassilenko, Ricardo Afonso, and many more.

I also want to thank Dra. Manuela Pereira from the Burn Unit in Santa Maria Hospital, Lisbon, for the wonderful guided visit to their installations. It allowed me to have a better view of the reality of clinical burn injury treatment.

Of course, I cannot forget my classmates. I had the privilege to interact with three different years during the course. I met some really interesting and smart people which became friends. Many have potential to become the future of healthcare innovation in Portugal. I need to leave a special note to Hélio Ochoa. He introduced me to the robotic system and the control laws we used. He was my main helper and without him I would not have accomplished so much. Thank you for being there when I needed.

Last but not least, my friends and family. To all my friends who put up with my ramblings about science and technology, and to those who saw me not so often, thank you for the support and understanding. To my mom, thank you for putting me into this world and raising me. To my brother, thanks for the support and your 3D printer ;). Finally, to my wife and child, I want to apologise for all the hours stolen. I hope to repay them with interest. Thank you for the unconditional support and being there on the dark times.\\

Much love to all.

