%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter9.tex
%% NOVA thesis document file
%%
%% Chapter with the system validation, tests and results.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{System Validation}
\label{cha:system_validation}

\begin{quotation}
\begin{flushright}
\itshape
«In theory there is no difference between theory and practice.\\ But in practice, there is.»\\
\textbf{- Benjamin Brewster}
\end{flushright}
\end{quotation}

The present chapter describes the system validation tests setup. Which tests are conducted, what they consist in, and why are they important. It follows with the exposition of the test results both on simulation and real environments. At the end, the results obtained are discussed and compared with current state of the art.

% ==========================
% = Physical Description =
% ==========================

\section{Validation Tests Setup}
\label{sec:system_validation_tests_setup}

The system developed on this thesis was tested on two different environments. First it was tested in a simulated environment to validate the concept and prevent any type of hardware damage. After accomplishing the desired goals in simulation, the system was also tested on a real environment.\\

In general, the test setup was the same for both environments. There were two levels of testing. The first level is testing at module level. Each component's layers were tested individually. The second level is integration testing where the whole system was tested to accomplish the goal of semi-autonomous wound filling 3D bioprinting.

The test list is the following:

\begin{itemize}
    \item Wound detection & segmentation
    \item Spatial data processing
\end{itemize}

% section system_validation_tests_setup

% ==========================
% = Simulation Results =
% ==========================

\section{Simulated System Results}
\label{sec:system_validation_simulation_results}

% section system_validation_simulation_results

% ==========================
% = Real Results =
% ==========================

\section{Real System Results}
\label{sec:system_validation_real_results}

% section system_validation_real_results

% ==========================
% = Discussion =
% ==========================

\section{Discussion}
\label{sec:system_validation_discussion}

% section system_validation_discussion